Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-06-02 submission/Sze_Obesity_PeerJ_2016.Rmd
Edit discussion section

---
title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: msystems.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

## Abstract
Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity. This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets. The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis. Alpha-diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances. Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%. A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity. When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`). Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.

@@

***Data Analysis***
We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness. After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e. BMI > 35.0). We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package. For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences. Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha-diversity metrics were below or above the median value for the study. The relative risk (RR) was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group. We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 within each study and across all of the studies using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages. Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models. For each study we developed models using either OTUs or genus-level phylotypes. The quality of each model was assessed by measuring the area under the curve (AUC) of the Receive Operating Characteristic (ROC) using ten-fold cross validation. Because the genus-level phylotype models were developed using a common reference, it was possible to use one study's model (i.e. the training set) to classify the samples from the other studies (i.e. the testing sets). The optimum threshold for the training set was set as the probability threshold that had the highest combined sensitivity and specificity. This threshold was then used to calculate the accuracy of the model applied to the test studies. To generate Receiver Operator Characteristic (ROC) curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package. Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions. We also calculated the actual sample size needed based on the effect size of each individual study.

@@

***Development of a microbiome-based classifier of obesity***
The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The the observed AUC values for the models applied to the training datastes varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this to identify a threshold to calculate the accuracy of the models when applied to the other number_string[n-1] datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

@@

Our meta-analysis helps to provide a little more clarity to the on going debate of whether or not there are specific microbiome variables that are influenced by obesity. Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that the data do not support that there is a microbiome difference between non-obese and obese individuals. However, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give. Here we perform an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. We analyzed the data using three different approaches and also performed post-hoc power and sample size calculations and simulations. The first approach uses the method taken by the previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014], the second approach takes a meta-analysis approach using RR representation by forest plots and funnel plots, the final approach uses linear mixed effect models to adjust for study effects and invetigates both predicted power and sample number needed for the observed effect sizes.

Overall, the first approach in this study by and large agrees with the prevous two reviews[@walters_meta-analyses_2014; @finucane_taxonomic_2014], in that for the majority of studies, no significant correlation between the microbiome and obesityy could be observed. Further, even though random forest could identify specific variables with similar taxonomic classifications that resulted in good ROC curves [Figure 1] these variables varied widely by both their actual sequence and directionality.  Similarily, for the B:F ratio we found that the other two approaches, that used the pooled data approach, were also not able to identify a correlation between measures of the microbiome and obesity [Figure 3 & 4]. For these approaches both the B:F ratio failed to have a significant correlation to obesity.  However, a small but significant differences was observed between non-obese and obese individauls  for Shannon diversity [Figure 3 & 4].  In order to also assess the amount of our results that could be due to insufficient power we performed power and sample size analysis.  Overall, we found that most studies for Shannon Diversity would be powered to find a 15% difference, however, the vast majority of reported effect sizes between studies ranged from `r format(round(min(alpha_effectSize[, 'shannon']), 2), nsmall=2)`-`r format(round(max(alpha_effectSize[, 'shannon']), 2), nsmall=2)` percent.  Although the effect sizes for the BF ratio were larger than that for Shannon Diversity [Figure 5A & 6A] the estimated power remained low.  Additionaly, we would also need a relatively large sample size to be able to detect the actual differences in each study for both Shannon Diversity and B:F ratio [Figure 5B & 6B].  Taken together we believe that this analysis builds on the previous two re-anlaysis and provides additional information as to whether there is a difference between non-obese and obese human individuals with respect to both Shannon diversity and the B:F ratio.

We do acknowledge that this study does have a few limitations that are worth mentioning. First,  it has been documented that in a number of groups BMI is not a good classification of obesity [@who_expert_consultation_appropriate_2004; @rahman_accuracy_2010]. Of particular note, the asian population has a separate set of guidelines for obesity [@who_expert_consultation_appropriate_2004]. Although the asain population in the studies analyzed are very small [Table 1] it is possible that this may contribute to some of the noise in the data. However, it is important to keep in mind that for these measures, specifically the RR, there does not seem to be any study bias due to studies having more ethnic diversity [Figure 4]. Additionally, we are also limited in our BMI classifications by what the previous studies published and performed. In general, not enough studies have been done to investigate the microbiome in the context of other markers for obesity or on different ethnicities to be able to perform the analysis that we did. Thus as more specific studies are published that address ethnicity and studies on the microbiome that use better markers of obesity are completed it is entirely possible that the noise associated with these measurements will decrease.

Second, although we make attempts to correct for study effects in our pooled analysis  so that we can compare the microbiome across studies it is possible that the different variable regions sampled, sequencing machines used, and the quality of the output data could have an impact on the overall conclusions reached.  To help mitigate some of these problems we analyzed each data set separately and processed all samples through a uniform pipeline in an attempt to mitigate a portion of the variation that might occur due to sequence processing.  It should also be noted that we also do not see any specific bias in the forest plots [Figure 3] based on hypervariable region.  Further many of the studies show very similar actual effect sizes for Shannon Diversity difference between non-obese and obese individuals.  However, even with all these precautions and safety checks we do realize that it may not be possible to compare these studies together and that further studies need to be completed to verify the observations made in this report.

Third, although we may not have detected a signal between non-obese and obese individuals our results indicate that we may not have been sufficiently powered to do so. However, taking both the sample size calculations and power calculations together would indicate that if there is a true difference to be found between non-obese and obese individuals this difference would be very small.  Additionally, due to the very small effect size potentially the more important question that does need to be raised is whether or not this difference is truly biologically meaningful.

Fourth, although the ROC curves for each study were relatively good there was very little actual overlap between studies based on sequence similarity and directionality.  Further none of the selected variables from each indivdiual study did a respectable job of classification when applied to the other data sets.  This would suggest that within each indivdiual study classification of non-obese and obese is relatively strong but not generalizable across studies.  This might indicate an ancestry, ethnicity, or a specific environmental effect such as diet difference between study populations that might cause this dissimilarity.  Additionally, an alternative explanation that may be more reasonable is that the random forest model is overfit for some of the studies examined which explains why the variables selected are less generalizable across studies.  In actuality it may be a combination of all of the factors mentioned as to why generalizability is problematic.

Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity.  There is a large body of very strong evidence in mice on the effect the microbiome has on obesity [@turnbaugh_effect_2009; @koren_host_2012].  It is not possible to discount these findings and it may be the case that in humans there is just too many other variables and factors that effect the microbiome which makes the signal too hard to detect in a cross sectional manner.  Some well designed studies in humans and diet, in the past, have shown that the noise inherent to microbiome studies can be overcome with very careful attention made to controlling the study conditions [@ley_microbial_2006].

Overall, this study provides a reasonable follow up to the two previous re-analysis studies to assess the association between the microbiome and obesity. In doing so we believe that we have provided evidence that if there is an effect size present it is small and will take many more studies and samples to be analyzed to confirm a difference between non-obese and obese individuals.

## Acknowledgements

The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the manuscript. Finally, we would like to thank all the authors who have made their data publicly available.  Without this initial forethought studies such as these would not be possible.

>>>>

---
title: "Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: msystems.csl
fontsize: 12pt
geometry: margin=1.0in
---

@@

## Abstract
Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity. This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets. The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis. Alpha diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances. Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%. A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity. When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]` (median=`r median(test_summary$accuracy)`). Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.

@@

***Data Analysis***
We split the overall meta-analysis into three general strategies using R (`r paste0(R.Version()$major,'.', R.Version()$minor)`). First, we followed the approach employed by Finucane et al [-@finucane_taxonomic_2014] and Walters et al [-@walters_meta-analyses_2014] where each study was re-analyzed separately to identify associations between BMI and the relative abundance of Bacteroidetes and Firmicutes, the ratio of Bacteroidetes and Firmicutes relative abundances (B:F), Shannon diversity, observed richness, and Shannon evenness. After each variable was transformed to fit a normal distribution a two-tailed t-test was performed for comparison of non-obese and obese individuals (i.e. BMI > 35.0). We performed a pooled analysis on these measured variables using linear random effect models to correct for study effect to asses differences on the combined dataset between non-obese and obese groups using the lme4 (v.`r packageDescription("lme4")$Version`) R package. Next, we compared the community structure from non-obese and obese individuals using PERMANOVA analysis of Bray-Curtis distance matrices. This analysis was performed using the vegan (v.`r packageDescription("vegan")$Version`) R package. For both analyses, the datasets were rarefied (N=1000) so that each study within a study had the same number of sequences. Second, for each study we partitioned the subjects into a low or high group depending on whether their alpha diversity metrics were below or above the median value for the study. The relative risk (RR) was then calculated as the ratio of the number of obese individuals in the low group to the number of obese individuals in the high group. We then performed a Fisher exact-test to investigate whether the RR was significantly different from 1.0 within each study and across all of the studies using the epiR (`r packageDescription("epiR")$Version`) and metafor (`r packageDescription("metafor")$Version`) packages. Third, we used the AUCRF (`r packageDescription("AUCRF")$Version`) R package to generate Random Forest models. For each study we developed models using either OTUs or genus-level phylotypes. The quality of each model was assessed by measuring the area under the curve (AUC) of the Receive Operating Characteristic (ROC) using ten-fold cross validation. Because the genus-level phylotype models were developed using a common reference, it was possible to use one study's model (i.e. the training set) to classify the samples from the other studies (i.e. the testing sets). The optimum threshold for the training set was set as the probability threshold that had the highest combined sensitivity and specificity. This threshold was then used to calculate the accuracy of the model applied to the test studies. To generate Receiver Operator Characteristic (ROC) curves and calculate the accuracy of the models we used the pROC (`r packageDescription("pROC")$Version`) R package. Finally, we performed power and sample number simulations for different effect sizes for each study using the pwr (`r packageDescription("pwr")$Version`) R package and base R functions. We also calculated the actual sample size needed based on the effect size of each individual study.

@@

***Development of a microbiome-based classifier of obesity***
The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The the observed AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this to identify a threshold to calculate the accuracy of the models when applied to the other number_string[n-1] datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

@@

Our meta-analysis helps to provide a clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity. We performed an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics. We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome is limited. Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes. Considering these datasets are among the largest published, it appears that most microbiome studies are underpowered to detect differences in alpha diversity.

Alpha diversity metrics are attractive because they distill a complex dataset to a single value. For example, diversity is a measure of the entropy in a community and integrates richness and evenness information. Two communities with little taxonomic similarity can have the same diversity. Among ecologists the relevance of these metrics is questioned because it is difficult to ascribe a mechanistic interpretation to their relationship with stability or disease. Regardless, the concept of a biologically significant effect size needs to be developed among microbiome researchers. Alternative metrics could include the ability to detect a defined difference in the relative abundance of an OTU representing a defined relative abundance. What makes for a biologically significant difference or relative abundance is an important point that has yet to be discussed in the microbiome field. The use of operationally defined effect sizes should be adequate until it is possible to decide upon an accepted practice.

By selecting a range of possible effect sizes, we were able to demonstrate that most studies are underpowered to detect modest differences in alpha diversity metrics and phylum-level relative abundances. Several factors interact to limit the power of microbiome studies. There is wide interpersonal variation in the diversity and structure of the human microbiome. In addition, the common experimental designs limit their power. As we observed, most of the studies included in our analysis were unbalanced for the variable that we were interested in. This was also true of those studies that originally sought to identify associations with obesity. Even with a balanced design, we showed that it was necessary to obtain approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)` and `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)` sequences per sample to detect a 5% difference in Shannon diversity or B:F, respectively. It was interesting that these sample sizes agreed across studies regardless of their sequencing method, region with in the 16S rRNA gene, or subject population (Figure 6). This suggests that regardless of the treatment or category, these sample sizes represent a good starting point for subject recruitment when using stool samples. Unfortunately, few studies have been published with this level of subject recruitment. This is troubling since the positive predictive rate of a significant finding in an underpowered study is small leading to results that cannot be reproduced [ref]. Future microbiome studies should articulate the basis for their experimental design.

Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference. Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from 3 studies [refs] and a metagenomic study and the Walters study used 16S rRNA gene sequence data from 5 studies [refs]; two studies were included in both studies. Our analysis included 4 of these studies [refs] and excluded 3 of the studies because they were too small [ref], only utilized metagenomic data [ref], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering. The additional six datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies. Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study. To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available. Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g. subject's sex or age).

Similar to our study, the Walters analysis [@walters_meta-analyses_2014], the authors generated Random Forest machine learning models to differentiate between non-obese and obese individuals. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.

Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity. There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005]. These studies appear to conflict with the observations using human subjects. Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity. Alternatively, it could be that the involvement of the microbiome in obesity is at the level of a common set of metabolites that can be produced from different structures of the microbiome.

## Acknowledgements
The authors would like to thank Nielson Baxter and Shawn Whitefield for their suggestions on the development of the manuscript. We are grateful to the authors of the studies used in our meta-analysis who have made their data publicly available or available to us directly. Without their forethought studies such as this would not be possible. This work was supported in part by funding from the National Institutes of Health to PDS (U01AI2425501 and P30DK034933).
